---
title: "SPY Vol Regime + Momentum Strategy"
author: "Dante Peinado"
date: "`r Sys.Date()`"
output: html_document
---
## Overview

This project implements a systematic long-only strategy on SPY combining two uncorrelated signals: a rolling volatility regime classifier and an EMA momentum filter. The strategy was developed in R using tidyquant and validated with a train/test split across 15 years of data (2010-2024).

The vol regime classifier architecture was inspired by dynamic threshold monitoring work done during an actuarial internship at Chubb Insurance, where I developed a sliding time-window percentile threshold for seasonal claims analysis. The same methodology is applied here to classify market environments.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(quantmod)
library(tidyquant)
library(TTR)

getSymbols("SPY", from = "2010-01-01", to = "2024-12-31")
```


## Data & Signal Construction

### Price Data & Momentum Signals

```{r spy-prices}
spy_prices <- data.frame(
  date = index(SPY),
  SPY.Adjusted = coredata(Ad(SPY))
) |>
  mutate(
    ema_fast = EMA(SPY.Adjusted, n=10),
    ema_slow = EMA(SPY.Adjusted, n=30)
  )
```

### Rolling Volatility & Regime Classification

The vol regime classifier uses a 10-day rolling annualized volatility estimate, compared against the dynamic 25th and 75th percentile thresholds calculated over a 252-day rolling window (full trading year). This produces three regime labels: high, neutral, and low.

```{r vol-regime}
spy_returns <- dailyReturn(SPY, type = "log")
roll_sd <- runSD(spy_returns, n = 10)
roll_ann_vol <- roll_sd * sqrt(252)

roll_vol_df <- data.frame(
  date = index(roll_ann_vol),
  roll_vol = coredata(roll_ann_vol)
)

roll_75 <- rollapply(roll_ann_vol, width = 252, FUN = quantile, probs = 0.75, na.rm = TRUE, align = "right")

roll_25 <- rollapply(roll_ann_vol, width = 252, FUN = quantile,probs = 0.25, na.rm = TRUE, align = "right")

regime <- ifelse(roll_ann_vol > roll_75, "high",
        ifelse(roll_ann_vol < roll_25, "low", "neutral"))

regime_df       <- data.frame(date = roll_vol_df$date, regime = regime)
regime_df_clean <- regime_df[!is.na(regime_df$regime), ]

```

## Visualization

## Vol Regime Chart

```{r regime-plot, fig.width=10, fig.height=5}
regime_rle <- rle(regime_df_clean$regime)
end_idx <- cumsum(regime_rle$lengths)
start_idx <- c(1, head(end_idx, -1) + 1)

regime_blocks <- data.frame(
  regime = regime_rle$values,
  xmin = regime_df_clean$date[start_idx],
  xmax = regime_df_clean$date[end_idx]
)

ggplot(roll_vol_df) +
  geom_rect(data = regime_blocks,
            aes(xmin = xmin, xmax = xmax,
                ymin = -Inf, ymax = Inf, fill = regime), alpha = 0.3) + geom_line(aes(x = date, y = roll_vol)) + labs(title = "SPY Rolling Annualized Volatility with Regime Classification",x = "Date", y = "Annualized Vol") + theme_minimal()
```

## STRATEGY

## Entry & Exit Rules

The strategy uses two conditions for entry and a looser OR condition for exit:

-**Entry**: Low vol regime and EMA(10) > EMA(30), a calm market with an upward momentum

-**EXIT**: High vol regime or EMA(10) < EMA(30), either danger signal triggers an exit.

This asymmetric structure of a strict entry and a loose exit is designed to protect capital by getting out quickly when either signal deteriorates.

### Backtesting Function

```{r backtest-function}
run_backtest <- function(data) {
  trades <- data.frame(
    entry_date = as.Date(character()),
    exit_date = as.Date(character()),
    direction         = as.character(),
    entry_price       = as.numeric(),
    exit_price        = as.numeric(),
    position_size     = as.numeric(),
    transaction_costs = as.numeric()
  )
  
  in_position <- FALSE
  entry_date  <- NULL
  entry_price <- NULL
  
  for (i in 1:nrow(data)) {
    if (!in_position & data$regime[i] == "low" & data$ema_fast[i] > data$ema_slow[i]) {
      entry_date  <- data$date[i]
      entry_price <- data$SPY.Adjusted[i]
      in_position <- TRUE
    } else if (in_position & (data$regime[i] == "high" | data$ema_fast[i]< data$ema_slow[i])) {
      trades <- rbind(trades, data.frame(
        entry_date = entry_date,
        exit_date = data$date[i],
        direction = "long",
        entry_price = entry_price,
        exit_price = data$SPY.Adjusted[i],
        position_size = 1,
        transaction_costs = 0.001
        
      ))
      in_position <- FALSE
      entry_date  <- NULL
      entry_price <- NULL
    }
  }
  
  trades |>
    mutate(PnL = (exit_price - entry_price) * position_size - transaction_costs)
}

```

## Results

```{r results}
dates <- inner_join(spy_prices, roll_vol_df, by = "date") |>
  inner_join(regime_df_clean, by = "date")

train <- filter(dates, date < as.Date("2021-01-01"))
test <- filter(dates, date >= as.Date("2021-01-01"))

trades_train <- run_backtest(train)
trades_test  <- run_backtest(test)

summarize_trades <- function(trades, label)
{
  cum_pnl      <- cumsum(trades$PnL)
  running_peak <- cummax(cum_pnl)
  
  data.frame(
    period = label,
    total_pnl = round(sum(trades$PnL), 2),
    win_rate = round(sum(trades$PnL > 0) / nrow(trades) * 100, 1),
    avg_pnl = round(mean(trades$PnL), 2),
    sharpe = round(mean(trades$PnL) / sd(trades$PnL), 2),
    max_dd = round(min(cum_pnl - running_peak), 2),
    n_trades = nrow(trades)
  )
}

results <- rbind(
  summarize_trades(trades_train, "Train (2010-2020)"), summarize_trades(trades_test, "Test (2021-2024)")
)

knitr::kable(results, caption = "Strategy Performance Summary")
```

## Walk-Forward Validation
To assess strategy viability across different market regimes, a rolling 
walk-forward test was conducted using 3-year training windows and 1-year 
test windows from 2013 to 2024.

```{r walk-forward}

wf_results <- data.frame()

for (test_year in 2013:2024) { 
  train_start <- as.Date(paste0(test_year - 3, "-01-01"))  
  train_end   <- as.Date(paste0(test_year, "-01-01")) - 1  
  test_start  <- as.Date(paste0(test_year, "-01-01")) 
  test_end    <- as.Date(paste0(test_year + 1, "-01-01"))
  
  train_data    <- filter(dates, date >= train_start & date < test_start)
  test_data     <- filter(dates, date >= test_start & date < test_end)
  train_results <- run_backtest(train_data)
  test_results  <- run_backtest(test_data)
  
  wf_results <- rbind(wf_results, summarize_trades(test_results, as.character(test_year)))
}

knitr::kable(wf_results, caption = "Walk-Forward Test Results by Year")

```

The strategy performs well in trending low-vol environments (2013, 2021, 2023, 2024) 
but struggles in choppy or declining markets (2015-2016, 2018-2020) where regime 
signals whipsaw frequently. This motivates the addition of an ADX trend strength 
filter as a next iteration.



## Limitations & Next Steps

- Position sizing is fixed at 1 share: next iterations will implement Kelly Criterion based sizing
- Strategy is Long-only: short signals via momentum reversal to be explored
- Currently limited to SPY: generalization to sector ETFs and individual equities planned.
- Walk-forward testing reveals whipsaw weakness in choppy markets
- ADX trend strength filter planned as next iteration

